---
title: "EDA calcio"
output: html_document
date: '2022-03-09'
---

# Setup

```{r}
# getwd() to identify working directory
```

Read the CSV file

```{r}
calcio <- read.table("data/calcio.csv", header = TRUE, sep = ",")
```

```{r}
dim(calcio)
```

```{r}
head(calcio)
```
```{r}
tail(calcio)
```

To sum up, this dataset contains seven variables. Each row is a football match of the "Serie A", the italian primary league, from 2008 to 2015.

The variables "Date", "HomeTeam", and "AwayTeam" are self-explanatory.

The variable "FTR" means "Full TIme Result" and it indicates whether the match has been won by the home team, the away team, or if it ended in a draw.

Finally, the variables "B365H", "B365D", and "B365A" indicate the odds of the betting company named "Bet365 Group Ltd" relative to the three events, obviously before the start of the match.

# Data Manipulation

Now we can observe and access the columns' names.

```{r}
colnames(calcio)
```

Let's explore the variables types.

```{r}
str(calcio)
```

Quantitative variables are codified in R as 'integer' (discrete) or 'numeric' (continuous).

Qualitative variables are codified in R as 'character' (text string) or 'factor' (repeated values).

Dates are codified in R as 'Date'.

```{r}
print(is.numeric(calcio$B365H))
print(is.numeric(calcio$Date))
```

```{r}
class(calcio$B365D)
```

```{r}
# first elements in a vector
calcio$FTR[1:10]
```

We need to convert the following variables to factor.

```{r}
calcio$HomeTeam <- factor(calcio$HomeTeam)
calcio$AwayTeam <- factor(calcio$AwayTeam)
calcio$FTR <- factor(calcio$FTR)
```

Alternatively, we could've added the option "stringsAsFactors = TRUE" inside the read.table function.

```{r}
class(calcio$HomeTeam)
```

```{r}
# to access the modalities of a factor variable
levels(calcio$HomeTeam)
```

```{r}
# to rename modalities
# levels(calcio$FTR) <- c("Away", "Draw", "Home")
```

```{r}
calcio$Date <- as.Date(calcio$Date, format = "%Y-%m-%d")
class(calcio$Date)
```

```{r}
print(min(calcio$Date))
print(max(calcio$Date))
```

Select only draws in a new df.

```{r}
calcio_draw <- calcio[calcio$FTR == "D", ]
head(calcio_draw)
```

Select the matches where the Home team has a low probability of winning the match.

```{r}
calcio_homeLow <- calcio[calcio$B365H > 9, ]
calcio_homeLow
```

Detect missing values.

```{r}
calcio[rowSums(is.na(calcio)) > 0, ]
```

The above function tells us how many values are missing in each row.

```{r}
# to remove missing values from df
# calcio_noNA <- na.omit(calcio)
```

Subsetting for the matches where the Home team has a low probability of winning the match.

```{r}
calcio_homeLow <- subset(calcio, subset = B365H > 9)
calcio_homeLow
```

Subset remove missing values. Subset can select rows and/or columns.

# Bonus

In the soccer bets the quotas represent the inverse of the probability of a event, for which

$(Probability Of The Event) = 1 / (Share Of The Event)$

The quota also represents the amount in € that the winner receives against the payment of the unit amount of € 1.

If the bet were fair, the odds of winning, drawing and losing would have to add up to 1.

However, this does not happen: the odds obtained add up to a value > 1.

This discrepancy is called bookmaking or premium and guarantees the bank a profit. (in italian "allibramento" o "aggio")

```{r}
calcio$premium <- 1 / calcio$B365H + 1 / calcio$B365D + 1 / calcio$B365A - 1
```

```{r}
subset(calcio, HomeTeam == "Milan")
```

```{r}
calcio[which.min(calcio$premium), ]
calcio[which.max(calcio$premium), ]
```

```{r}
calcio09_10 <- subset(calcio, Date <= "2009-08-15" & Date <= "2010-06-15")
calcio09_10[which.max(calcio09_10$premium), ]
```